---
title: "The *ToothGrowth* data analysis."
author: "Maciej Nowak"
date: "Tuesday, July 21, 2015"
output: pdf_document
---

##Exploratory data analysis
We will be working on the *ToothGrowth* data that comes with R. The R documentation titles this data as *"The Effect of Vitamin C on Tooth Growth in Guinea Pigs"* and that Vitamin C was served as orange juice (OJ) or ascorbic acid (VC - "pure" Vitamin C).
First we load the data:
```{r load, echo=TRUE}
rm(list = ls()) # clear the environment so the results are reproducible
data(ToothGrowth) # load the data
class(ToothGrowth) # check the class/data type
```
Now that we know it is a *data.frame* we can continue our analysis:
```{r, echo=TRUE}
head(ToothGrowth, 3) # check the columns and how the data looks like
nrow(ToothGrowth) # how many rows are there?
levels(ToothGrowth$supp) # how many supplement types are there? 
unique(ToothGrowth$dose) # is the dose discrete or continuous?
ToothGrowth[ which.max(ToothGrowth$len), ] # the record with max len
ToothGrowth[ which.min(ToothGrowth$len), ] # the record with min len
which(is.na(ToothGrowth$len)) # any NA len
```
With this information we can draw a graph to illustrate how consumption of orange juice or vitamin C and the doses impact teeth growth. Let us first explore the data layout:
```{r, echo=TRUE}
library(ggplot2)
g <- ggplot(ToothGrowth, aes(x = dose, y = len))
g <- g + geom_point() +
     facet_wrap(~ supp) + 
     labs(x = "Dose [mg]") +
     labs(y = "Tooth length [mm]") +
     labs(title = "Tooth length for OJ (Orange Juice) and Vitamin C") 
g
```

Then we would check the data distribution by *supp* and *dose*. The distribution for orange juice is in the top row and for VC in the bottom row:
```{r, echo=TRUE}
g <- ggplot(ToothGrowth)
g <- g + geom_histogram(aes(x = len)) + facet_wrap(supp ~ dose) +
     labs(x = "Tooth length [mm]") +
     labs(title = "Tooth length distribution for OJ and Vitamin C")
g
```

##Comparison of tooth growth by supp and dose
So we want to determine whether consumption of orange juice rather than vitamin C would increase teeth growth and whether it depends on the dose. We split *ToothGrowth* by doses and calculate the respective confidence intervals:  
```{r, echo=TRUE}
doses <- unique(ToothGrowth$dose)
```
**The confidence interval for `r doses[ 1 ]`mg**
```{r, echo=TRUE}
d <- ToothGrowth[ ToothGrowth$dose == doses[ 1 ], ]
t.test(len ~ supp, data = d, paired = FALSE, var.equal = TRUE)
```
**The confidence interval for `r doses[ 2 ]`mg**
```{r, echo=TRUE}
d <- ToothGrowth[ ToothGrowth$dose == doses[ 2 ], ]
t.test(len ~ supp, data = d, paired = FALSE, var.equal = TRUE)
```
**The confidence interval for `r doses[ 3 ]`mg**
```{r, echo=TRUE}
d <- ToothGrowth[ ToothGrowth$dose == doses[ 3 ], ]
t.test(len ~ supp, data = d, paired = FALSE, var.equal = TRUE)
```
From the results printed I would say that for doses `r doses[ 1 ]` and `r doses[ 2 ]` orange juice seems to help growing longer teeth. We cannot say this for the `r doses[ 3 ]` dose since the confidence interval  spreads across zero rather symmetrically and the mean in the OJ group and in the VC group are almost the same.