---
title: "Statistical Inference Course Project - The Exponential Distribution"
author: "Maciej Nowak"
date: "Monday, July 20, 2015"
output: pdf_document
---

##Overview
This project concentrates on the law of large numbers and the central limit theorem both illustrated by the exponential distribution with rate parameter set here to 0.2. We will run 1000 simulations and show that the mean and variance approach their theoretical values for adequately large number of simulations. The first step is to set up the driving variables.
```{r setup, echo=TRUE}
rm(list = ls()) # clear the environment so the results are reproducible
lambda <- 0.2 # set the rate parameter 
n <- 40 # exponentials' count
simCount <- 1000 # simulations' count
```

##Sample Mean versus Theoretical Mean
Here the theoretical mean is `r 1 / lambda` (1/`r lambda`), we will generate a vector (*means*) of `r simCount` means of `r n` exponentials then cumulate them and calculate the cumulative averages. Then we plot the sample means and the theoretical mean. As we can see both lines converge as the sample count grows.
```{r mean, echo=TRUE}
theoreticalMean <- 1 / lambda
means <- c()
for(i in 1:simCount)
{
    means <- c(means, mean(rexp(n, lambda)))
}
m <- cumsum(means) / (1:simCount)

plot(x = 1:simCount,
     y = m,
     type = "l",
     main = "The sample mean versus the theoretical mean",
     xlab = "Simulations",
     ylab = "Mean",
     col = "black")

lines(x = c(1, simCount),
      y = c(theoreticalMean, theoreticalMean),
      type = "l",
      col = "red")

legend("topright", pch = "_",
       col = c("black", "red"),
       legend = c("sample mean", "theoretical mean"))
```

##Sample Variance versus Theoretical Variance
Here the theoretical variance is `r 1 / lambda^2` (1/`r lambda^2`), we will generate a vector (*vars*) of `r simCount` variances of `r n` exponentials then cumulate them and calculate the cumulative averages. Then we plot the sample variance and the theoretical one. As we can see both lines converge as the sample count grows.
```{r variance, echo=TRUE}
theoreticalVar <- 1 / lambda^2
vars <- c()
for(i in 1:simCount)
{
    vars <- c(vars, var(rexp(n, lambda)))
}
v <- cumsum(vars) / (1:simCount)

plot(x = 1:simCount,
     y = v,
     type = "l",
     main = "The sample variance versus the theoretical one",
     xlab = "Simulations",
     ylab = "variance",
     col = "black")

lines(x = c(1, simCount),
      y = c(theoreticalVar, theoreticalVar),
      type = "l",
      col = "red")

legend("bottomright", pch = "_",
       col = c("black", "red"),
       legend = c("sample variance", "theoretical variance"))
```

##Distribution
Finally we want to show that with accordance to the central limit theorem as the sample size increases the distribution becomes that of a standard normal. We use the previously generated mean vector (*means*) to draw a histogram that resembles the standard normal one, as we can see the distribution "concentrates" around the theoretical mean of `r 1 / lambda` (1/`r lambda`).
```{r distribution, echo=TRUE}
hist(means,
     main = paste("Distribution of", simCount, "means"),
     xlab = "Mean")
```
